Psychological Needs During Intergroup Contact: Three Experience Sampling Studies

### 2022-11-02 manuscript.tex
Updates from Overleaf

% Abstract
\abstract{
\hl{needs to be adjusted}\\
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. This question becomes especially relevant when considering that negative intergroup contacts are common during everyday life. To understand positive intergroup interactions, we propose to consider situational motivations and psychological needs during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total N of measurements = 10,297). Across the three studies, we find inconsistent results for a traditional test of the contact hypothesis with experience sampling data. However, when using multilevel models to test the contact hypothesis, we find that individual outgroup interactions and perceived interaction quality have a strong influence on positive outgroup attitudes. 
The fulfillment of psychological needs is consistently a strong predictor of positive attitudes during intergroup interactions and these positive effects emerged via perceived interaction quality. The situational needs remained the core predictor of outgroup attitudes even after controlling for Allport’s (\citeyear{Allport1954b}) contact conditions or more general self-determination needs \citep[][]{Ryan2017}. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

% Migrant Example Pathways
% Relevance (Version 1 of 2): Migrant Example Version:
Conflict between social groups and their individual members remains a prevalent feature of the modern human condition. One of the main examples to date are the struggles of many migrants across the world, hoping to build a positive relationship with the majority group. One of the main ameliorations proposed by social psychologists has been the intergroup contact hypothesis. In its most essential interpretation, the intergroup contact hypothesis postulates that prejudice can be reduced and favorable attitudes increased if members of two groups have frequent and positive contact \citep[e.g.,][]{Allport1954b, Hewstone1996, Pettigrew1998}. Over the past 70 years, a plethora of studies and interventions have shown the general effectiveness of positive intergroup contact \citep[e.g.,][]{Pettigrew2006}. However, even though a central assumption of intergroup contact theory has been that the contact should be positive, relatively little research has thus far explained when and why people perceive their everyday inter-group interactions as positive.

>>>>

% Abstract
\abstract{
\hl{needs to be adjusted}\\
One challenge of modern intergroup contact research has been the question of when and why an interaction is perceived as positive and leads to better intergroup relations. We propose to consider situational psychological needs during everyday intergroup contact. We conducted three extensive longitudinal studies with recent migrants, to capture their interactions with the majority outgroup (total N of measurements = 10,297). \hl{We find inconsistent results for a traditional test of the contact hypothesis with experience sampling data. However, when using multilevel models to test the contact hypothesis, we find that individual outgroup interactions and perceived interaction quality have a strong influence on positive outgroup attitudes.} 
The fulfillment of psychological needs is consistently a strong predictor of positive attitudes during intergroup interactions and these positive effects emerged via perceived interaction quality. The situational needs remained the core predictor of outgroup attitudes even after controlling for Allport’s (\citeyear{Allport1954b}) contact conditions or more general self-determination needs \citep[][]{Ryan2017}. As one of the first studies to test intergroup contact theory using extensive longitudinal data, we offer insight into the mechanisms of positive intergroup contact during real-life interactions and find situational motivations to be a key building block of understanding and addressing positive intergroup interactions.

@@

% Migrant Example Pathways
% Relevance (Version 1 of 2): Migrant Example Version:
Conflict between social groups and their individual members remains a prevalent feature of the modern human condition. One of the main examples to date are the struggles of many migrants across the world, hoping to build a positive relationship with the majority group. One of the main ameliorations proposed has been intergroup contact. In its most essential interpretation, the intergroup contact hypothesis postulates that prejudice can be reduced and favorable attitudes increased if members of two groups have frequent and positive contact \citep[e.g.,][]{Allport1954b, Hewstone1996, Pettigrew1998}. Over the past 70 years, a plethora of studies and interventions have shown the general effectiveness of positive intergroup contact \citep[e.g.,][]{Pettigrew2006}. However, even though a central assumption of intergroup contact theory is that the contact should be positive, relatively little research has thus far explained when and why people perceive their everyday inter-group interactions as positive.
